---
title: 'WGCNA Pipeline - part two: Selected lipids based on maternal diet'
author: "Rachel Richardson"
output:
  html_document: default
  pdf_document: default
---

###Adapted from the ANOVAs used for plotting interaction and comparing chain length, this code pulls out specific lipid species with different abundances in groups from the metadata. Because we have so many lipids that can be compared to our gene clusters, this step helps us determine which lipids would be the most likely related to our biological group of interest. For our data, we are intrested in the effects of maternal diet.

```{r package install, echo = FALSE, eval= FALSE}
install.packages("exactRankTests")
install.packages("ggplot2")
install.packages("reshape2")
install.packages("ggsignif")

```


###**From scratch, we would need to load our data to use (if you have not already done so in previous steps).** 

```{r Data import, echo = T}


#Read in lipid abundance data to be compared:

DGs1.2 <- read.csv("1.2DGs.changed LOQ.csv", head = TRUE, row.names = 1)
AC <- read.csv("AC.formatted.noloq.csv", head = TRUE, row.names = 1)
Cer <- read.csv("Cerimides.changedLOQ.csv", head = TRUE, row.names = 1)

DGs1.3 <- read.csv("CM033018 1-3DGs.changedLOQ.csv", head = TRUE, row.names = 1)
names(DGs1.3) <- paste(names(DGs1.3), ".1.3DGs", sep = "") #Correction for poor naming

dh <- read.csv("dhCer.changedLOQ.csv", head = TRUE, row.names = 1)
Glu <- read.csv("GluCer.changedLOQ.csv", head = TRUE, row.names = 1)
hex <- read.csv("hexosylCer.changedLOQ.csv", head = TRUE, row.names = 1)
Lac <- read.csv("LacCer.changedLOQ.csv", head = TRUE, row.names = 1)
mye <- read.csv("Sphingomyelins.formatted.noloq.csv", head = TRUE, row.names = 1)
sine <- read.csv("Sphingosine.formatted.noloq.csv", header = TRUE, row.names = 1)
TAG <- read.csv("TAG.changedLOQ.csv", head = TRUE, row.names = 1)
```


###Our lipid abundance dataset originated from several excel sheets across different lipid species, which we reformatted as needed and saved as CSVs. We also took into account if any data points fell below the limit of quantification (LOQ) and created seperate csvs with adjusted values based on LOQ (noted as changed LOQ) or noted that data points were unmodified since all points were within range of the experiment (no LOQ).

###**Collected metadata contains the information for different experimental groups across samples.** However, metadata was collected on a larger group of subjects than the ones tested for lipid abundance, therefore we filter the data by the our subjects of interest.


```{r metadata import, echo = T}

#Read in metadata, location of groupings to be tested:

  #Selects for subjects with lipid abundance samples, as based on sine dataset
metadata <- read.table("metadata.txt", header = TRUE)[row.names(sine),]     

  #Creates a new column with all groupings of interest as a factor. Format:  MaternalBodyMass.MaternalDiet-PostweanDiet
metadata$combination <- as.factor(paste(metadata$M_BM, paste(metadata$M_Diet,metadata$PW_Diet, sep = "-"), sep = "."))  

```

###The resulting lipid data CSVs are arranged with different lipid species as columns and different subjects as rows. The metadata is formatted similarly, where columns are different categorical variables and rows are subjects.

```{r show format, echo = T, warning=FALSE}
library(knitr)

#Lipid data format
kable(head(AC[1:7]))
#Metadata format
kable(head(metadata))

```

###In order to determine the effect of diet, we are most interested in the lipid species with significantly different abundances between the groups. One common way would be to run ANOVA or t-test using the groups notated in the metadata. 

###However, ANOVA testing assumes normality of the residuals in our data, which can be approximated by overall distribution when sample size is not small. We know from our previous distribution analyses that the raw data does not appear to be normally distributed, but we can double-check our eye-ball rejection of normality with a Shapiro-Wilk (test of normality where the null hypothesis is that samples come from a normal distribution).


```{r normal data values, echo= TRUE}


#Select p-values for testing purposes.

shapvalue <- 0.05

#Contain all data sets in a list as variable and a string list

datanames <- list(DGs1.2, AC, Cer, DGs1.3, dh, Glu, hex, mye, Lac, sine, TAG)
datanamesstr <- c('DGs1.2', 'AC', 'Cer', 'DGs1.3', 'dh', 'Glu', 'hex', 'mye', 'Lac', 'sine', 'TAG')

#Apply Shapiro-Wilk across datasets

totalprint <- c()   # Vector to hold outcomes across datasets
iter <- 1           # iterable for which dataset is being processed

for (datan in datanames){ #For each lipid dataset read in,
  testresult <- array()            #create a new array for holding normality test results.
  
  for(x in 1:length(datan)){                            #For every column in that dataset, 
    test <- shapiro.test(unlist(datan[,x]))     #perform an ANOVA based on the groups specified in the metadata,
    testresult[x] <- test$p.value                 #and save the isolated significance value from the anova output in the sig array.
  }
  
                                                   
  total <- length(testresult)                                 # Counts all species in processed dataset
  rejectnull <- length(testresult[testresult < shapvalue])    # Counts all species not from a normal distribution in a processed dataset 
  
  totalprint <- c(totalprint,                                 # Adds to totalprint vector
                  sprintf("From the %s dataset, %d/%d reject the null hypothesis that samples are drawn from a normal distribution.", 
                  datanamesstr[iter], rejectnull,total))
  iter <- iter + 1
}

totalprint
```


###Similarly, we can check if transformations are roughly sufficient. We used a log10 transformation here:


```{r Adjusted data values, echo= TRUE}
#Select p-values for testing purposes.

shapvalue <- 0.05

#Contain all data sets in a list as variable and a string list

datanames <- list(DGs1.2, AC, Cer, DGs1.3, dh, Glu, hex, mye, Lac, sine, TAG)
datanamesstr <- c('DGs1.2', 'AC', 'Cer', 'DGs1.3', 'dh', 'Glu', 'hex', 'mye', 'Lac', 'sine', 'TAG')

#Apply Shapiro-Wilk across datasets

totalprint <- c()   # Vector to hold outcomes across datasets
iter <- 1           # iterable for which dataset is being processed

for (datan in datanames){ #For each lipid dataset read in,
  testresult <- array()            #create a new array for holding normality test results.
  
  for(x in 1:length(datan)){                            #For every column in that dataset, 

    #######    
    
    test <- shapiro.test(unlist(log10(datan[,x])))     #perform an ANOVA based on the groups specified in the metadata, use this line for transformation changes
    
    #######
    
    testresult[x] <- test$p.value                 #and save the isolated significance value from the anova output in the sig array.
  }
  
                                                   
  total <- length(testresult)                                 # Counts all species in processed dataset
  rejectnull <- length(testresult[testresult < shapvalue])    # Counts all species not from a normal distribution in a processed dataset 
  
  totalprint <- c(totalprint,                                 # Adds to totalprint vector
                  sprintf("From the %s dataset, %d/%d reject the null hypothesis that samples are drawn from a normal distribution.", 
                  datanamesstr[iter], rejectnull,total))
  iter <- iter + 1
}

totalprint

```


### Our raw data does not seem to meet the assumptions of ANOVA due to non-normality and our log-transform data seems questionable. **An alternative method to the ANOVA that doesn't assume normality is the Mann-Whitney-Wilcoxon Test, which we will use here to analyze differences based on maternal diet and post-wean diet. Because Mann-Whitney-Wilcoxon Test only accepts two levels for grouping factors, we use Kruskal-Wallis test (also non-parametric comparison of distributions) to test the combined factor that includes maternal body mass, maternal diet, and post-wean diet (five levels, one for each experimental group).** 

####Note: Significance rejects the null hypothisis, where both datasets are sampled from the same distribution

```{r library exactrank, echo= TRUE, results='hide', warning=FALSE, message=FALSE}
library(exactRankTests)
```

```{r non-parametric across comparisons of interest, echo= TRUE}
#Contain all data sets in a list

datanames <- list(DGs1.2, AC, Cer, DGs1.3, dh, Glu, hex, mye, Lac, sine, TAG)

#Define significance p-value; selects for lipids of interest

pvalue <- 0.05

################################
#All combonation Kruskal-Wallis


# New vector to hold anova lipid species and if abundances are significantly different among all five groups
ALL <- c() 


for (datan in datanames){ #For each lipid dataset read in,
  sig <- array()            #create a new array for holding significance values.
  
  for(x in 1:length(datan)){                            #For every column in that dataset, 
    kru <- kruskal.test(datan[,x]~unlist(metadata[,6]))     #perform test based on the groups specified in the metadata,
    sig[x] <- kru$p.value                                    #and save the isolated significance value from the anova output in the sig array.
  }
  
  #Sig array will have all significance values for each lipid species in the dataset
  names(sig) <- colnames(datan)     #Label sig values with the column names of the dataset
  ALL <- c(ALL, sig[sig < pvalue])    #Add to vector, ALL, that is a subset of all sig arrays with significance less than defined p-value
}

#Edit column names and row names for clarity; specific to our dataset
  #Rename column as "Significance"
ALL <- data.frame(ALL)
colnames(ALL) <-  "Significance"

 #Global substitute for row name format
rnames <- gsub("..pmol.", "", row.names(ALL))
rnames <- gsub("X", "", rnames)
row.names(ALL) <- rnames


#View data format
kable(head(ALL))

#######################

#Maternal diet Wilcox test

Mat <- c() # New vector to hold anova lipid species and if abundances are significantly different among all five groups

for (datan in datanames){ #For each lipid dataset read in,
  sig <- array()            #create a new array for holding significance values.
  
  for(x in 1:length(datan)){                            #For every column in that dataset, 
    wil <- wilcox.exact(datan[,x]~unlist(metadata[,3]))     #perform an wilcox test based on the groups specified in the metadata,
    sig[x] <- wil$p.value                                  #and save the isolated significance value from the anova output in the sig array.
  }
  
  #Sig array will have all significance values for each lipid species in the dataset
  names(sig) <- colnames(datan)     #Label sig values with the column names of the dataset
  Mat <- c(Mat, sig[sig < pvalue])     #Add to vector, Mat, that is a subset of all sig arrays with significance less than defined p-value
}

#Edit column names and row names for clarity; specific to our dataset
  #Rename column as "Significance"
Mat <- data.frame(Mat)
colnames(Mat) <-  "Significance"

  #Global substitute for row name format
rnames <- gsub("..pmol.", "", row.names(Mat))
rnames <- gsub("X", "", rnames)
row.names(Mat) <- rnames

#View data format
kable(head(Mat))

#######################

#Offspring diet Wilcox test

PW <- c()

for (datan in datanames){ #For each lipid dataset read in,
  sig <- array()            #create a new array for holding significance values.
  
  for(x in 1:length(datan)){                            #For every column in that dataset, 
    wil <- wilcox.exact(datan[,x]~unlist(metadata[,4]))     #perform an wilcox test based on the groups specified in the metadata,
    sig[x] <- wil$p.value                                  #and save the isolated significance value from the anova output in the sig array.
  }
  
  #Sig array will have all significance values for each lipid species in the dataset
  names(sig) <- colnames(datan)     #Label sig values with the column names of the dataset
  PW <- c(PW, sig[sig < pvalue])     #Add to vector, PW, that is a subset of all sig arrays with significance less than defined p-value
}
  
#Edit column names and row names for clarity; specific to our dataset
  #Rename column as "Significance"
PW <- data.frame(PW)
colnames(PW) <-  "Significance"

  #Global substitute for row name format
rnames <- gsub("..pmol.", "", row.names(PW))
rnames <- gsub("X", "", rnames)
row.names(PW) <- rnames


#View data format
kable(head(PW))

#Save as method specific; we will be looking at an ANOVA/t-test example with similar code as well

ALL.K <- ALL
Mat.W <- Mat
PW.W <- PW


```



###**If our transformed samples would meet normality the assumptions, below is an example of how you would use ANOVA and t-test to analyze differences based on maternal diet, post-wean diet, and a combined factor that includes maternal body mass, maternal diet, and post-wean diet.** 

```{r ANOVA across comparisons of interest, echo= TRUE}

#Define significance p-value; selects for lipids of interest

pvalue <- 0.05

################################
#All combonation ANOVA

# New vector to hold anova lipid species and if abundances are significantly different among all five groups
ALL <- c() 


for (datan in datanames){ #For each lipid dataset read in,
  sig <- array()            #create a new array for holding ANOVA significance values.
  
  for(x in 1:length(datan)){                            #For every column in that dataset, 
    ano <- anova(aov(log10(datan[,x])~unlist(metadata[,6])))     #perform an ANOVA based on the groups specified in the metadata,
    sig[x] <- as.double(ano$`Pr(>F)`[1])                  #and save the isolated significance value from the anova output in the sig array.
  }
  
  #Sig array will have all significance values for each lipid species in the dataset
  names(sig) <- colnames(datan)     #Label sig values with the column names of the dataset
  ALL <- c(ALL, sig[sig < pvalue])    #Add to vector, ALL, that is a subset of all sig arrays with significance less than defined p-value
}

#Edit column names and row names for clarity; specific to our dataset
  #Rename column as "Significance"
ALL <- data.frame(ALL)
colnames(ALL) <-  "Significance"

 #Global substitute for row name format
rnames <- gsub("..pmol.", "", row.names(ALL))
rnames <- gsub("X", "", rnames)
row.names(ALL) <- rnames


#View data format
kable(head(ALL))

#######################

#Maternal diet t-test

Mat <- c() # New vector to hold anova lipid species and if abundances are significantly different among all five groups

for (datan in datanames){ #For each lipid dataset read in,
  sig <- array()            #create a new array for holding ANOVA significance values.
    
  for(x in 1:length(datan)){                            #For every column in that dataset, 
    ttest <- t.test(log10(datan[,x])~unlist(metadata[,3]))     #perform test based on the groups specified in the metadata,
    sig[x] <- ttest$p.value                              #and save the isolated significance value from the anova output in the sig array.
  }
  
  #Sig array will have all significance values for each lipid species in the dataset
  names(sig) <- colnames(datan)     #Label sig values with the column names of the dataset
  Mat <- c(Mat, sig[sig < pvalue])     #Add to vector, Mat, that is a subset of all sig arrays with significance less than defined p-value
}

#Edit column names and row names for clarity; specific to our dataset
  #Rename column as "Significance"
Mat <- data.frame(Mat)
colnames(Mat) <-  "Significance"

  #Global substitute for row name format
rnames <- gsub("..pmol.", "", row.names(Mat))
rnames <- gsub("X", "", rnames)
row.names(Mat) <- rnames

#View data format
kable(head(Mat))

#######################

#Offspring diet t-test

PW <- c()

for (datan in datanames){ #For each lipid dataset read in,
  sig <- array()            #create a new array for holding ANOVA significance values.
  
  for(x in 1:length(datan)){                            #For every column in that dataset, 
    ttest <- t.test(log10(datan[,x])~unlist(metadata[,4]))     #perform test based on the groups specified in the metadata,
    sig[x] <- ttest$p.value                            #and save the isolated significance value from the anova output in the sig array.
  }
  
  #Sig array will have all significance values for each lipid species in the dataset
  names(sig) <- colnames(datan)     #Label sig values with the column names of the dataset
  PW <- c(PW, sig[sig < pvalue])     #Add to vector, PW, that is a subset of all sig arrays with significance less than defined p-value
}
  
#Edit column names and row names for clarity; specific to our dataset
  #Rename column as "Significance"
PW <- data.frame(PW)
colnames(PW) <-  "Significance"

  #Global substitute for row name format
rnames <- gsub("..pmol.", "", row.names(PW))
rnames <- gsub("X", "", rnames)
row.names(PW) <- rnames


#View data format
kable(head(PW))

#Save as method specific for comparison to non-parametric

ALL.A <- ALL
Mat.T <- Mat
PW.T <- PW

```


###**Comparing the two methods, we can see that there are different lipids identified as significant. We can also tell which lipids are identified as significant by both methods, a more conservative approach for finding lipids of interest.**

```{r Comparison of significance, echo= TRUE, evaluate = FALSE}

#Generate booleans for indexes of lipids shared between lists and use to find overlapping lipids of significance

AllAinK <- row.names(ALL.A) %in% row.names(ALL.K)
AllKinA <- row.names(ALL.K) %in% row.names(ALL.A)

cat("ALL overlap:  \n", row.names(ALL.A)[AllAinK[AllKinA]]) #All shared genes between two methods

PWTinW <- row.names(PW.T) %in% row.names(PW.W)
PWWinT <- row.names(PW.W) %in% row.names(PW.T)


cat("PW overlap: \n", row.names(PW.T)[PWTinW[PWWinT]]) #All shared genes between two methods

#Comparison of interest for maternal diet; conveniently the overlap between both methods is equal to Mat.W
#i.e., all lipids identified as significant in the Wilcoxon test were also significant by t-test.

MatTinW <- row.names(Mat.T) %in% row.names(Mat.W)
MatWinT <- row.names(Mat.W) %in% row.names(Mat.T)


cat("Indexes of Mat.T in Mat.W:  \n", which(MatTinW), "\n")
cat("Indexes of Mat.W in Mat.T:  \n", which(MatWinT), "\n")
cat("Mat.W:   \n", row.names(Mat.W))

```

###**Since in our comparison of interest (Maternal Diet) all non-parametric significant lipids are also in the list of parametric significant lipids, we will use this list going forward in our WGCNA analysis.**

####Note: We also evaluated for differences between sexes (not shown here), resulting in only two significant lipid species (16.0.20.4.1.2DGs, 0.04930763;			24.0dhCer, 0.02698315). Evaluation for maternal bodymass can also be tested, however results would not be conclusive given the imbalence of obese and lean maternal subjects.

###**For a easily communicated graphic, a barchart can adequately display abundance comparisons between groups. Here, we use ggplot2, reshape2, and ggsignif on a log-scale for visual clarity.**

```{r library plotting, echo= TRUE, eval=TRUE, results='hide', warning=FALSE}

library(ggplot2)
library(reshape2)
library(ggsignif)
```


```{r plot data, echo= TRUE}


Alllipid <- data.frame(DGs1.2, AC, Cer, DGs1.3, dh, Glu, hex, mye, Lac, sine, TAG)

 #Global substitute for column name format, keeps consistancy with tests for significance
cnames <- gsub("..pmol.", "", colnames(Alllipid))
cnames <- gsub("X", "", cnames)
colnames(Alllipid) <- cnames


#Pull out columns containing lipids of interest based on significance tests
Intlipid <- Alllipid[,row.names(Mat.W)]
Intlipid$M_Diet <- metadata[,3] #Add grouping of interest from significance test to dataframe


CTR <- subset(Intlipid, M_Diet=="CTR") #Subset of CTR maternal group
HFD <- subset(Intlipid, M_Diet=="HFD") #Subset of HFD maternal group

CTR <- CTR[1:length(CTR)-1] #Remove maternal diet from subset data
HFD <- HFD[1:length(HFD)-1] #Remove maternal diet from subset data


#Create a dataframe with the means of each lipid species subset (HFD, CTR)
plotter <- cbind(lapply(CTR, mean), lapply(HFD, mean))
plotter <- cbind(plotter, row.names(plotter))
plotter <- data.frame(unlist(plotter[,1]), unlist(plotter[,2]), row.names(plotter))
colnames(plotter)<-c("CTR", "HFD", "Lipid species")

#View Format
kable(head(plotter))


#Create a dataframe with the standard deviation of each lipid species subset (HFD, CTR)
plottersd <- cbind(lapply(CTR, sd), lapply(HFD, sd))
plottersd <- cbind(plottersd, row.names(plottersd))
plottersd <- data.frame(unlist(plottersd[,1]), unlist(plottersd[,2]), row.names(plottersd))
colnames(plottersd)<-c("CTR", "HFD", "Lipid species")


#Create a dataframe with the standard error of each lipid species subset (HFD, CTR)
#Standard error, standard deviation divided by square root of sample size

plotterse <- data.frame(plotter$`Lipid species`)
plotterse$CTR <- plottersd$CTR/sqrt((nrow(CTR)))
plotterse$HFD <- plottersd$HFD/sqrt((nrow(HFD)))
colnames(plotterse)<-c("Lipid species", "CTR", "HFD")

#View format
kable(head(plotterse))


#Melt into appropriate data structure for ggplot2
plotterm <- melt(plotter[,c("Lipid species","CTR", "HFD")],id.vars = 1)
plotterse <- melt(plotterse[,c("Lipid species","CTR", "HFD")],id.vars = 1)

plotterm$se <- plotterse$value #Add SE to original plotter for ease of graphing

#View format
kable(head(plotterm))


#Plot data

plot <- ggplot(plotterm,aes(x =`Lipid species`, y = value)) + 
    geom_bar(aes(fill = variable),stat = "identity",position = "dodge") + 
  
    #Use log scale and allow for room considering error and significance bars
    scale_y_log10(limits = c(1, 100000)) +
  
    #Light theme
    theme_light() +
  
    #Adjust text size and angle
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 14),
        axis.text.y = element_text(size = 14), axis.title.x = element_text(size = 16), axis.title.y = element_text(size = 16),
        plot.title = element_text(size = 16)) +
  
    #Adjust labels
    labs(title = "Significant lipid differences based on maternal diet", x = "Lipid Species", y = "Lipid abundance (pmol)", fill = "Maternal Diet") +
  
    #Add error bars based on standard error
    geom_errorbar(aes(ymin=value-se, ymax=value+se), position = position_dodge2(width=.2))
    
#Add and position significance bars
plot1 <- plot + geom_signif(y_position=4.5, 
                  xmin=c(seq(from = 0.8, to = length(row.names(plotter))-0.2, by =1 )), 
                  xmax=c(seq(from = 1.2, to = length(row.names(plotter))+0.2, by =1 )),
                
                  # Set astrisks; * = <0.05 (based on initial cutoff), ** = < 0.01, *** = < 0.001, etc
                  annotation=c(strrep("*", floor(log10(1/Mat.W$Significance)))), tip_length=0) 

plot1
```


###Save your plot if you'd like and save your lipids of interest to use in WGCNA! (Below code saves as an R object)

```{r saving data, echo= TRUE}

#Readjusting for size and height
png("MatSigbarplot.png", height = 900, width = 900)
plot +    
  
#Adjust text size and angle
theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 26),
        axis.text.y = element_text(size = 26), axis.title.x = element_text(size = 30), axis.title.y = element_text(size = 30),
        plot.title = element_text(color = "white"), legend.text = element_text(size = 26), legend.title = element_text(size = 30)) +  #Changed to blank title

  
#Add and position significance bars
geom_signif(y_position=4.5, 
              xmin=c(seq(from = 0.8, to = length(row.names(plotter))-0.2, by =1 )), 
              xmax=c(seq(from = 1.2, to = length(row.names(plotter))+0.2, by =1 )),
                
              # Set astrisks; * = <0.05 (based on initial cutoff), ** = < 0.01, *** = < 0.001, etc
              annotation=c(strrep("*", floor(log10(1/Mat.W$Significance)))), tip_length=0, textsize = 10) 

dev.off()

# Save an object to a file
saveRDS(Mat.W, file = "Mat.W.rds")


```
